# Setting up your workspace: Rstudio Projects

First, we'll set up an R project, a method for managing your work within RStudio. RStudio projects allow you to keep all folders and files associated within a given project together. The project will automatically control your working directory. 

To do this:
Create a new R studio Project: File -> New Project -> choose directory and project name

### Loading necessary "packages" for managine files and cleaning data

Load the `here` package and `tidyverse` package in your script to help with working directory and file paths. We'll use this a little. 

If you don't have them, you'll have to install them.

```{r eval=FALSE}
install.packages("tidyverse")
install.packages("here")


```


To load the necessary packages so you can use them, you'll have to run the commands below in each new R session.
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)

# check your working directory

here()

```


## Setting up the working directory 

To make life easier, we'll follow a general file/working directory structure. There are many ways to set up a working directory, but a simple and easy way to do this involves creating files for your data (sometimes with a subdirectory or new directory for cleaned or altered data), the scripts you'll use for running analysis, and the resultant output and plot. 

So, in the same directory (aka, folder) as your new RStudio project:

1. Create a folder called `scripts`
2. Create a folder called `data`
3. Create a folder called `output`
4. Create a folder called `plots`

