# Person Abilities
Person abilities are also of interest. We can look at the person side of the model by computing person abilities. 

1. Compute person abilities using the `tam.wle` function and assign to an object called `abil`. 
2. Extract person abilities ($\theta_p$) from `abil` to create an object in the `environment` called `PersonAbility` which will essentially be a column vector. 

**Note**: You may want more information than this at times (such as standard errors) so you may not always want to subset this way.

```{r}

#generates a data frame - output related to estimation
abil <- tam.wle(mod1)
```
See the first few rows of Abil. Notice you get:

1. `pid`: person id assigned by TAM.
2. `N.items`: Number of items the person was given (this becomes interesting when you have linked test forms where students may not all see the same number of items)
3. `PersonScores`: Number of items the student got right or endorsed (in the survey case).
4. `PersonMax`: Max total that person could have gotten right/selected an option for
5. `theta`: estimated person ability
6. `error`: estimated measurement error
7. `WLE.rel`: estimated person seperation reliability.

```
head(Abil)

# or

View(Abil)
```
```{r echo=FALSE}
kable(head(Abil))
```


The column in the `abil` data.frame corresponding to person estimates is the `theta` column. Pull out the ability estimates, theta, column if you would like, though, this creates a list. This makes it a little easier for a few basic tasks below.
```{r}

PersonAbility <- abil$theta

```
```{r include=FALSE}
PersonAbility[1:20]
```

```{r}
# Only the first 6 rows, shown
head(PersonAbility)
```

You can export those estimated abilites to a .csv to save (you can also save directly in R, if you need to). This writes `abil` as a csv file to your `output` directory that we created earlier using the `here` package. 

```
write.csv(abil, here("output", "HLSmod1_thetas.csv")
```

Descriptives for person ability
```{r}
hist(PersonAbility)
mean(PersonAbility)
sd(PersonAbility)
```

## Wright Map
To visualize the relationship between item difficulty and person ability distributions, call the WrightMap package installed previously. We'll generate a simple WrightMap. We'll clean it up a little bit by removing some elements

```{r}
library(WrightMap)
IRT.WrightMap(mod1)
IRT.WrightMap(mod1, show.thr.lab=FALSE)
```

### Exercise: 
1. Are the items appropriately targeted to the ability level of the population? 
2. Why do you think?